Double Standards in Judging Collective Action

### 2023-05-31 Documents/manuscript.qmd
submit

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias in terms of class, race, and gender. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In three preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions---but do not show consistent ingroup bias (in terms of class, race, and gender) when judging collective action by ingroup and outgroup members.

@@

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable (normative) and unacceptable (non-normative) forms of collective action. For that purpose, we define a double standard as judging the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting. [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.]

@@

Partisan differences can result from two social psychological processes. On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological orientations shape how we think and feel about issues, actions, and events. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depend more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In practice, the two processes are often difficult to disentangle.

@@

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more acceptable if the protesters' are ingroup members or the protesters' cause aligns with the ingroup's interests (*Hypothesis 1a*, *1b*) or if the protesters' cause aligns with their own system-justifying or system-challenging motives (*Hypotheses 2a*, *2b*). In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*, *2b*). Building on Experiments 1--2, Experiment 3 examines how liberal and conservative men and women judge collective action for or against restricting abortion in anticipation of the Supreme Court overturning its *Roe v. Wade* decision on reproductive rights in the United States. By varying the protesters' cause as well as the participants' ideology and gender, Experiment 3 compares the relative evidence for identity-based (*Hypothesis 1b*) and ideology-based (*Hypothesis 2b*) double standards in judging collective action concerning a conservative cause in a sample balanced by gender and ideology. Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action.

@@

Model 1 (M1) estimated varying intercepts for the eight conditions to test whether, in line with Hypotheses 1a, 1b, or 2a, participants' responses depended on their own group membership, the protesters' group membership, or the protesters' cause. Contradicting *Hypothesis 1a*, participants did not consider protest actions performed by their ingroup to be more acceptable, on average, than the same actions performed by the relevant outgroup (`r with(m1, d_text[contrast == "h1a_contrast"])`). Contradicting *Hypothesis 1b*, participants did not consider protest actions for a cause that was nominally aligned with their ingroup's interests to be more acceptable, on average, than protest actions for a cause not aligned with their ingroup's interests (`r with(m1, d_text[contrast == "h1b_contrast"])`). Contradicting *Hypothesis 2a*, participants did not consider protest actions for a system-defending cause to be more acceptable, on average, than protest actions for a system-challenging cause (`r with(m1, d_text[contrast == "h2a_contrast"])`). Instead, we found that, on average, White participants considered all protest actions to be more acceptable than Black participants (`r with(m1, d_text[contrast == "nh1_contrast"])`) and both Black (`r with(m1, d_text[contrast == "nh3_contrast"])`) and White (`r with(m1, d_text[contrast == "nh2_contrast"])`) participants considered the same actions to be more acceptable when protests were for, rather than against, defunding the police. As in Experiment 1, we thus found that participants' responses depended on both the participants' and the protesters' group memberships---but not in the directions predicted by our hypotheses.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a function of their *z*-standardized endorsement of system-justifying beliefs. As preregistered, we modeled this relationship with two fixed effects, one estimating the effect of system-justifying beliefs on judgments about system-defending protest actions and one estimating their effect on judgments about system-challenging protest actions, and one varying effect estimating its variance across conditions. Supporting *Hypothesis 2b*, participants who *rejected* system-justifying beliefs were *more* likely to consider system-challenging protest actions (for defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "for"])`). We did not, however, find evidence for the ideological symmetry implied by *Hypothesis 2b* since participants who *endorsed* system-justifying beliefs were *not* more likely to consider system-defending protest actions (against defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "against"])`).

@@

Model 3 (M3) extended Model 1 by estimating participants' responses as a function of their self-reported support for the cause of the protest. We recoded participants' responses to create a predictor variable that encoded support for defunding the police when protesters supported defunding the police and opposition to defunding the police when protesters opposed defunding the police. As preregistered, we modeled this relationship as a monotonic effect [@burkner_modelling_2020] that estimated the average change in the outcome variable across predictor categories as well as how much of this change occurred between each of the four pairs of adjacent predictor categories.^[Our model assigned a Dirichlet prior, $\alpha = {1, 1, 1, 1}$, to the proportions of the overall change that was expected to occur between each of the four pairs of predictor categories.] Contradicting *Hypothesis 3*, participants who supported the protesters' cause did not, on average, consider the same protest actions to be more acceptable than participants who opposed the protesters' cause (`r with(m3_h3, b_text)`). Ergo, our results did not support the alternative hypothesis that ideology-based double standards can be reduced to support for, or opposition to, the cause of a protest.

@@

  # Calculate contrasts
  m5 <- m5 %>% 
    group_by(.draw, x_ide) %>% 
    summarize(
      across(
        ends_with("_contrast"),
        ~mean(z * .)
      )
    ) %>% 
    pivot_longer(
      ends_with("_contrast"),
      names_to = "contrast",
      values_to = "d"
    ) %>% 
    group_by(x_ide, contrast) %>% 
    median_qi(d)

@@

\noindent Participants were then asked to select which position the text described and what arguments for that position were presented in the text. If they selected the wrong answer, they were instructed to reread the text and select the right answer. Participants then rated to what extent they support or oppose restricting abortion and to what extent their feelings about this issue are a moral conviction.

@@

\noindent As in the previous experiments, participants rated which of several protest actions they consider acceptable means to protest for or against restricting abortion. On the next pages, participants completed the remaining measures. On the final page, participants responded to two attention checks that assessed whether they paid attention to the experimental manipulation: "In this study, you read about some people's view on a policy change. What was the policy change?" and "In this study, you then read about some protesters. Were the protesters for or against restricting abortion?".

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggested that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender. We did not find evidence for those double standards even though we examined social categories that are central to most people's identities, even when we considered only highly identified group members, and even though participants' reactions to the manipulation showed that, at least in Experiments 1--2, they understood the protesters' causes in ingroup--outgroup terms.

@@

First, we might understand ideology-based double standards as an expression of ingroup bias based on opinion-based [@mcgarty_collective_2009] or partisan [@osborne_psychology_2022; @finkel_political_2020] identities. This explanation aligns with Verkuyten et al.'s [-@verkuyten_intolerance_2022] finding that participants were more tolerant of transgressive protest actions when taken by their most-liked, rather than their least-liked, political group (e.g., climate activists). An identity-based account, however, implies that both conservatives and progressives would show ingroup bias and, therefore, cannot explain the observed asymmetry in the two groups' double standards.

@@

Lastly, we might consider that collective action, even in its controversial forms, is not judged as a moral transgression [as in previous research on identity-based double standards, e.g., @abrams_double_2013]---but is instead understood as a means to an end and judged in relation to its end ("the end justifies the mean"). Their different moral concerns [@graham_liberals_2009; @kivikangas_moral_2021] might lead progressives and conservatives to see their support of certain causes as a fundamental matter of right or wrong [@skitka_psychology_2021] and, by thus moralizing those causes, to accept more extreme means to achieving them. This explanation aligns with Richardson & Conway's [-@richardson_standing_2022] finding that moral concerns explained why liberals rated protesting for liberal causes as more moral than conservatives (and vice versa). A values-based account could also explain why conservatives who, more so than liberals, endorse moral concerns related to loyalty and authority  tend to reject disruptive protest actions even when the protesters' cause otherwise aligns with their moral concerns.

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In five studies (total *N* = 2,979), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police and for or against restricting abortion in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own ideological position---but showed no consistent ingroup bias in terms of class, race, and gender. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
keywords: collective action, double standards, social identity theory, system justification theory, Black Lives Matter
statement: |
  This research shows that what people consider an acceptable *means* of protest depends on what the *end* of a protest is. That is, people tend to judge the *same* controversial protest actions to be more acceptable when the protesters' cause aligned with their own ideological positions. This finding helps explain the often divided response to recent protest movements.
word-count: "11,546 words"
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

In this paper, we examine double standards in judging collective action---that is, whether observers judge the same protest actions as more or less acceptable depending on their own and the protesters' group memberships (*identity-based double standards*) or on the protesters' cause and how it aligns with their own ideological positions (*ideology-based double standards*). In two studies, we first develop an instrument of 25 controversial protest actions, based in item response theory, to capture double standards in judging collective action. In three preregistered experiments, we then use this instrument to test for double standards in judging protest actions for workers' rights, for or against defunding the police, and for or against restricting abortion. We find evidence for ideology-based, but not for identity-based, double standards: People judge the same protest actions to be more acceptable when the protesters' cause aligns with their own ideological positions---but do not show consistent ingroup bias (in terms of class, race, and gender) when judging collective action by ingroup and outgroup members.

@@

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable (normative) and unacceptable (non-normative) forms of collective action. For that purpose, we define a double standard as judging the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.]

@@

Partisan differences can result from two social psychological processes. On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological differences shape how we think and feel about issues, actions, and events. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depend more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In practice, the two processes are often difficult to disentangle.

@@

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more acceptable if the protesters' are ingroup members or the protesters' cause aligns with the ingroup's interests (*Hypothesis 1a*, *1b*) or if the protesters' cause aligns with their own system-justifying or system-challenging motives (*Hypotheses 2a*, *2b*).

In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*, *2b*). Building on Experiments 1--2, Experiment 3 examines how liberal and conservative men and women judge collective action for or against restricting abortion in anticipation of the Supreme Court overturning its *Roe v. Wade* decision on reproductive rights in the United States. By varying the protesters' cause as well as the participants' ideology and gender, Experiment 3 compares the relative evidence for identity-based (*Hypothesis 1b*) and ideology-based (*Hypothesis 2b*) double standards in judging collective action concerning a conservative cause in a sample balanced by gender and ideology. Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action.

@@

Model 1 estimated varying intercepts for the eight conditions to test whether, in line with Hypotheses 1a, 1b, or 2a, participants' responses depended on their own group membership, the protesters' group membership, or the protesters' cause. Contradicting *Hypothesis 1a*, participants did not consider protest actions performed by their ingroup to be more acceptable, on average, than the same actions performed by the relevant outgroup (`r with(m1, d_text[contrast == "h1a_contrast"])`). Contradicting *Hypothesis 1b*, participants did not consider protest actions for a cause that was nominally aligned with their ingroup's interests to be more acceptable, on average, than protest actions for a cause not aligned with their ingroup's interests (`r with(m1, d_text[contrast == "h1b_contrast"])`). Contradicting *Hypothesis 2a*, participants did not consider protest actions for a system-defending cause to be more acceptable, on average, than protest actions for a system-challenging cause (`r with(m1, d_text[contrast == "h2a_contrast"])`). Instead, we found that, on average, White participants considered all protest actions to be more acceptable than Black participants (`r with(m1, d_text[contrast == "nh1_contrast"])`) and both Black (`r with(m1, d_text[contrast == "nh3_contrast"])`) and White (`r with(m1, d_text[contrast == "nh2_contrast"])`) participants considered the same actions to be more acceptable when protests were for, rather than against, defunding the police. As in Experiment 1, we thus found that participants' responses depended on both the participants' and the protesters' group memberships---but not in the directions predicted by our hypotheses.

@@

Model 2 extended Model 1 by estimating participants' responses as a function of their *z*-standardized endorsement of system-justifying beliefs. As preregistered, we modeled this relationship with two fixed effects, one estimating the effect of system-justifying beliefs on judgments about system-defending protest actions and one estimating their effect on judgments about system-challenging protest actions, and one varying effect estimating its variance across conditions. Supporting *Hypothesis 2b*, participants who *rejected* system-justifying beliefs were *more* likely to consider system-challenging protest actions (for defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "for"])`). We did not, however, find evidence for the ideological symmetry implied by *Hypothesis 2b* since participants who *endorsed* system-justifying beliefs were *not* more likely to consider system-defending protest actions (against defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "against"])`).

@@

Model 3 extended Model 1 by estimating participants' responses as a function of their self-reported support for the cause of the protest. We recoded participants' responses to create a predictor variable that encoded support for defunding the police when protesters supported defunding the police and opposition to defunding the police when protesters opposed defunding the police. As preregistered, we modeled this relationship as a monotonic effect [@burkner_modelling_2020] that estimated the average change in the outcome variable across predictor categories as well as how much of this change occurred between each of the four pairs of adjacent predictor categories.^[Our model assigned a Dirichlet prior, $\alpha = {1, 1, 1, 1}$, to the proportions of the overall change that was expected to occur between each of the four pairs of predictor categories.] Contradicting *Hypothesis 3*, participants who supported the protesters' cause did not, on average, consider the same protest actions to be more acceptable than participants who opposed the protesters' cause (`r with(m3_h3, b_text)`). Ergo, our results did not support the alternative hypothesis that ideology-based double standards can be reduced to support for, or opposition to, the cause of a protest.

@@

  # Calculate contrasts
  m5 <- m5 %>% 
    group_by(.draw, x_ide) %>% 
    summarize(
      across(
        ends_with("_contrast"),
        ~sum(z * .)
      )
    ) %>% 
    pivot_longer(
      ends_with("_contrast"),
      names_to = "contrast",
      values_to = "d"
    ) %>% 
    group_by(x_ide, contrast) %>% 
    median_qi(d)

@@

\noindent Participants were then asked to select which position the text described and what arguments for that position were presented in the text. If they selected the wrong answer, they were instructed to reread the text and select the right answer. Participants then rated to what extent they supported or opposed restricting abortion and to what extent their feelings about this issue were a moral conviction.

@@

\noindent As in the previous experiments, participants rated which of several protest actions they consider acceptable means to protest for or against restricting abortion. On the next pages, participants completed the remaining measures. On the final page, participants responded to two attention checks that assessed whether they had paid attention to the experimental manipulation: "In this study, you read about some people's view on a policy change. What was the policy change?" and "In this study, you then read about some protesters. Were the protesters for or against restricting abortion?".

@@

Growing evidence shows that, in many situations, observers apply a more lenient standard when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggested that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender. We did not find evidence for those double standards even though we examined social categories that are central to most people's identities, even when we considered only highly identified group members, and even though participants' reactions to the manipulation showed that, at least in Experiments 1--2, they understood the protesters' causes in ingroup--outgroup terms.

@@

First, we might understand ideology-based double standards as an expression of ingroup bias based on opinion-based [@mcgarty_collective_2009] or partisan [@osborne_psychology_2022; @finkel_political_2020] identities. This explanation aligns with Verkuyten et al.'s [-@verkuyten_intolerance_2022] finding that participants were more tolerant of transgressive protest actions when taken by their most-liked, rather than their least-liked, political group (e.g., climate activists). An identity-based account, however, implies that both conservatives and progressives would show ingroup bias and, therefore, cannot explain the observed asymmetry in the two groups' judgements.

@@

Lastly, we might consider that collective action, even in its controversial forms, is not judged as a moral transgression [as in previous research on identity-based double standards, e.g., @abrams_double_2013]---but is instead understood as a means to an end and judged in relation to its end ("the end justifies the means"). Their different moral concerns [@graham_liberals_2009; @kivikangas_moral_2021] might lead progressives and conservatives to see their support of certain causes as a fundamental matter of right or wrong [@skitka_psychology_2021] and, by thus moralizing those causes, to accept more extreme means to achieving them. This explanation aligns with Richardson and Conway's [-@richardson_standing_2022] finding that moral concerns explained why liberals rated protesting for liberal causes as more moral than conservatives (and vice versa). A values-based account could also explain why conservatives who, more so than liberals, endorse moral concerns related to loyalty and authority  tend to reject disruptive protest actions even when the protesters' cause otherwise aligns with their moral concerns.
